Estimating Planetary Habitability via Particle Swarm Optimization of CES Production Functions.

### 2018-01-19 docs/report/report.tex
Finished the report. Mailed.

The search for extra-terrestrial life~\cite{Seth1,Biosig} and potentially habitable extrasolar
planets~\cite{Kepler,Presto} has been an international venture since Frank Drake's attempt with Project Ozma in the
mid-20th century~\cite{Ozma}, demanding large investments in cost and effort. \textcite{Exo1} confirmed the first
exoplanet in 1991. This marked the start of a trend that has lasted 25 years and yielded over 3,700 confirmed
exoplanets. There have been attempts to assess the habitability of these planets and to assign a score based on their
similarity to Earth. Two such habitability scores are the Cobb-Douglas Habitability Score (CDHS)~\cite{Bora,Saha} and
the Constant Elasticity Earth Similarity Approach (CEESA) score. Estimating these scores involves maximizing a
production function while observing a set of constraints on the input variables.

Under most paradigms, maximizing a continuous function requires calculating a gradient. This may not always be feasible
for non-polynomial functions in high-dimensional search spaces. Further, subjecting the input variables to constraints,
as needed by CDHS and CEESA, are not always straightforward to represent within the model. This paper presents an
approach to Constrained Optimization (CO) using the swarm intelligence metaheuristic. \pso\ (PSO) is a method for
optimizing a continuous function that does away with the need for a gradient. It employs a large number of particles
that traverse the search space converging toward a global best solution encountered by at least one of the
particles~\cite{PSO1,PSO2}.

\pso\ is a distributed method that requires simple mathematical operators and short segments of code, making it a
lucrative solution where computational resources are at a premium. Its implementation is highly parallelizable. It
scales with the dimensionality of the search space. The standard PSO algorithm does not deal with constraints but
through variations in initializing and updating particles constraints are straightforward to represent and adhere to, as
seen in Section~\ref{subsec:copso}. \textcite{Poli1,Poli2} carried out extensive surveys on the applications of PSO
reporting uses in Communication Networks, Machine Learning, Design, Combinatorial Optimization and Modeling, among
others.

This paper demonstrates the applicability of \pso\ in estimating CDHS and CEESA scores of an exoplanet by maximizing
their respective production functions (discussed in Sections~\ref{subsec:cdhs} and~\ref{subsec:ceesa}). CDHS considers
the planet's Radius, Mass, Escape Velocity and Surface Temperature, while CEESA includes a fifth parameter, the Orbital
Eccentricity of the planet. The Exoplanets Catalog hosted by the Planetary Habitability Laboratory, UPR Arecibo records
these parameters for each exoplanet in Earth Units~\cite{PHL}. Section~\ref{sec:results} reports the performance of PSO
and discusses the distribution of the habitability scores of the exoplanets.

@@

\subsection{Cobb-Douglas Habitability Score}\label{subsec:cdhs}
Estimating the Cobb-Douglas Habitability Score (CDHS)~\cite{Bora} requires estimating an interior score \linebreak
($\mathit{CDHS}_i$) and a surface score ($\mathit{CDHS}_s$) by maximizing the following production functions,
\begin{subequations}
  \begin{alignat}{4}
    Y_i\ &=\ {CDHS}_i\ &=&\ R^\alpha.&D^\beta\,,\label{eq:cdhsi}\\
    Y_s\ &=\ {CDHS}_s\ &=&\ {V_e}^\gamma.&{T_s}^\delta\,,\label{eq:cdhss}
  \end{alignat}
\end{subequations}
where, $R$, $D$, $V_e$ and $T_s$ are density, radius, escape velocity and surface temperature respectively. $\alpha$,
$\beta$, $\gamma$ and $\delta$ are the elasticity coefficients subject to,
\begin{equation}
  0 < \alpha,\beta,\gamma,\delta < 1\,.
\end{equation}
Equations~\ref{eq:cdhsi} and~\ref{eq:cdhss} are convex under either Constant Returns to Scale (CRS) or Decreasing
Returns to Scale (DRS) marked by two constraints on the elasticity coefficients,
\begin{subequations}
  \begin{alignat}{3}
    \text{CRS:} & \quad\alpha+\beta = 1,&\quad\gamma+\delta = 1\,,\\
    \text{DRS:} & \quad\alpha+\beta < 1,&\quad\gamma+\delta < 1\,.
  \end{alignat}
\end{subequations}
The final CDHS is the convex combination of the interior and surface CDHS values as given by,
\begin{equation}
  Y\ =\ w_i.Y_i + w_s.Y_s\,.
\end{equation}

\subsection{Constant Elasticity Earth Similarity Approach}\label{subsec:ceesa}
% Cite what?
The Constant Elasticity Earth Similarity Approach (CEESA) uses the following production function to estimate the
habitability score of an exoplanet,
\begin{equation}\label{eq:ceesa}
  Y = {(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{\eta}{\rho}}\,,
\end{equation}
where, $E$ is the fifth parameter denoting Orbital Eccentricity. The value of $\rho$ lies within $0<\rho\leq 1$.
The coefficients ($r$, $d$, $t$, $v$ and $e$) are constrained by,
\begin{subequations}
  \begin{align}
      0 < r,d,t,v,e < 1\,,\\
      r+d+t+v+e = 1\,.
  \end{align}
\end{subequations}
The value of $\eta$ is constrained by the scale of production used,
\begin{subequations}
  \begin{alignat}{3}
    \text{CRS:} & \quad 0 < \eta \leq 1\,,\\
    \text{DRS:} & \quad \eta = 1\,.
  \end{alignat}
\end{subequations}

@@

% Cite: PSO paper.
Particle Swarm Optimization (PSO)~\cite{PSO1} is a biologically inspired metaheuristic for finding the global minima of
a function.  Traditionally designed for unconstrained inputs, it works by iteratively converging a population of
randomly initialized solutions, called particles, toward a globally optimal solution. Each particle in the population
keeps track of its current position and the best solution it has encountered so far, called $\pbest$. Each particle also
has an associated randomized velocity used to traverse the search space. The swarm keeps track of the overall best
solution, called $\gbest$. Each iteration of the swarm updates the velocity of the particle towards its $\pbest$ and the
$\gbest$ values.

@@

On each iteration, the algorithm updates the velocity and position of each particle. For each particle, it picks two
random numbers $u_g, u_p$ from a uniform distribution, $U(0,1)$ and updates the particle velocity as indicated in
line~\ref{algline:vup}. Here, $\omega$ is the inertial weight and $\lambda_g,\lambda_p$ are the global and particle
learning rates. If the new position of the particle corresponds to a better fit than its $\pbest$, the algorithm updates
$\pbest$ to the new position. Once the algorithm has updated all particles, it updates $\gbest$ to the new overall best
position. A suitable termination criteria for the swarm, under convex optimization, is  when the $\gbest$ position has
not changed by the end of the iteration.

@@

Under these guidelines, the following equations represent the CDHS estimation under CRS as a Constrained Optimization problem,
\begin{equation}\label{eq:cdhscrs}
  \begin{aligned}
    & \underset{\alpha,\beta,\gamma,\delta}{\text{minimize:}}
    & Y_i = {-}R^\alpha.D^\beta,\ Y_s = {-}{V_e}^\gamma.{T_s}^\delta\\
    & \text{subject to:}
    &  {-}\phi + \epsilon \leq 0,\quad\phi - 1 + \epsilon \leq 0\,, &\quad\quad \forall \phi\in\{\alpha,\beta,\gamma,\delta\}\\
    && (\alpha+\beta-1) - \tau \leq 0,\quad(\gamma+\delta-1) - \tau \leq 0\,,\\
    && (1-\alpha-\beta) - \tau \leq 0,\quad(1-\gamma-\delta) - \tau \leq 0\,.
  \end{aligned}
\end{equation}
Under DRS the last two constraints for $Y_i$ and $Y_s$ are replaced with,
\begin{equation}\label{eq:cdhsdrs}
  \begin{aligned}
    \alpha + \beta + \epsilon - 1 &\leq 0\,,\\
    \gamma + \delta + \epsilon - 1 &\leq 0\,.
  \end{aligned}
\end{equation}
The CEESA score estimation under DRS is represented as,
\begin{equation}\label{eq:ceesadrs}
  \begin{aligned}
    & \underset{r,d,t,v,e,\rho,\eta}{\text{minimize}}
    & Y = {-}{(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{\eta}{\rho}}\\
    & \text{subject to}
    &   {-}\phi + \epsilon \leq 0,\quad\phi - 1 + \epsilon \leq 0\,, &\quad\quad \forall \phi\in\{r,d,t,v,e,\eta\}\\
    && \rho - 1 \leq 0,\quad \rho - 1 + \epsilon \leq 0\,,\\
    && (r+d+t+v+e-1) - \tau \leq 0\,,\\
    && (1-r-d-t-v-e) - \tau \leq 0\,.
  \end{aligned}
\end{equation}
Under CRS there is no need for the parameter $\eta$ (since $\eta=1$). Thus, the objective function for the problem reduces to,
\begin{equation}\label{eq:ceesacrs}
  \underset{r,d,t,v,e,\rho}{\text{minimize}}\quad
  Y = {-}{(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{1}{\rho}}\,.
\end{equation}

@@

The data set used for estimating the Habitability Scores of exoplanets, described in Section~\ref{sec:habscore}, was the
Confirmed Exoplanets Catalog maintained by the Planetary Habitability Laboratory (PHL)~\cite{PHL}. The catalog records
observed and modeled parameters for exoplanets confirmed by the Extrasolar Planets Encyclopedia. Table~\ref{tab:param}
describes the parameters from the PHL Exoplanets Catalog (PHL-EC) used for the experiment. Since surface temperature and
eccentricity are not recorded in Earth Units, we normalized these values by dividing them with Earth's surface
temperature (\SI{288}{\kelvin}) and eccentricity (\num{0.017}). PHL-EC assumes an Eccentricity of 0 when unavailable.
The PHL-EC records empty values for planets whose surface temperature is not known. We chose to drop these records from
the experiment.

Our implementation used $n=25$ particles to traverse the search space, with learning rates $\lambda_g=0.8$ and
$\lambda_p=0.2$. It used an inertial weight of $\omega=0.6$ and upper and lower bounds $\pm1.0$. It used an error
threshold of $\epsilon=\num{1e-6}$ to convert strict inequalities to non-strict inequalities, and a tolerance of
$\tau=\num{1e-7}$ to transform an equality constraint to a pair of inequalities. Further implementation details are
discussed in Appendix~\ref{app:imp}.

@@

As for real time taken to converge, PSO took \SI{666.85}{\second} ($\approx\SI{11}{\minute}\ \SI{7}{\second}$) to
estimate the CDHS under CRS for \num{1683} exoplanets, at an average of \SI{198.11}{\milli\second} for each planet for
each part of the CDHS.\@ For CDHS under DRS, it took \SI{638.69}{\second} ($\approx\SI{10}{\minute}\ \SI{39}{\second}$)
at an average of \SI{189.75}{\milli\second} for each part of the CDHS.\@ The CEESA estimates, requiring one calculation,
took a little over half the total time to execute. Under DRS it took a total of \SI{370.86}{\second}
($\approx\SI{6}{\minute}\ \SI{11}{\second}$) at \SI{220.36}{\milli\second} per planet, while under CRS it took
\SI{356.92}{\second} ($\approx\SI{5}{\minute}\ \SI{57}{\second}$) at \SI{212.07}{\milli\second} per planet.

@@

\pso\ mainly draws its advantages from being easy to implement and highly parallelizable. The algorithms described in
Section~\ref{sec:pso} use simple operators and straightforward logic. What is especially noticeable is the lack of the
need for a gradient, allowing PSO to work in high dimensional search spaces with a large number of constraints.
Particles of the swarm, in most implementations operate independently during each iteration, their updates can occur
simultaneously and even asynchronously, yielding much faster execution times than those outlined in
Section~\ref{sec:results}. However, since strict inequalities and equality constraints are not exactly represented, the
resulting solution may not be as accurate as direct methods. Despite this, using PSO to calculate the habitability
scores is beneficial when the number of input parameters are large, which further increases the number of constraints,
resulting in a model too infeasible for traditional optimization methods.

@@

  \pagebreak
  \section{Ensuring Convergence}\label{app:con}

@@

  To prevent this from happening our implementation maintained a counter that was incremented very time $\gbest$ changed
  by a value less than the threshold it incremented the counter. If the counter reached \num{100}, the algorithm
  terminated and reported the current value of $\gbest$ as the global optima. If the value of $\gbest$ changed by more
  than the threshold, it reset the counter to $\num{0}$. Since the function is convex, and all points are initialized to
  feasible solutions (thus, ensuring that a feasible $\gbest$ exists) there was no need to restrict the total number of
  possible iterations.

  This explains the spike at \num{0} in Figure~\ref{fig:itercdcrs}. For every exoplanet, the implementation waits a
  \num{100} iterations for declaring convergence, thus overstating the total iterations by \num{100}. To adjust for this
  offset, we subtracted \num{100} from the total iterations before constructing the histogram plots in
  Figures~\ref{fig:itercdcrs},~\ref{fig:itercddrs},~\ref{fig:itercecrs} and~\ref{fig:itercedrs}. Therefore, although the
  graphs report iterations to convergence for a significant number of planets, only the global best has not shifted
  locations, but the particles have actually converged to a common point around a region within the threshold of the
  global best.

@@

  Matrices $P$ and $V$ represent current position and velocity of all particles in the swarm.
  \begin{alignat*}{2}
    P &= [\ p_{ij}\ &&\mid\ \forall i=1\dots n,\;\forall j=1\dots d\ ]\,,\\
    V &= [\ v_{ij}\ &&\mid\ \forall i=1\dots n,\;\forall j=1\dots d\ ]\,.
  \end{alignat*}
  $p_{ij}$ and $v_{ij}$ denote position and velocity of particle $i$ in dimension $d$. Matrices $B$ and $L$ represent
  the $\pbest$ and $\lbest$ values for the swarm, where,
  \begin{align*}
    B &= [\ {\pbest}_i\ \mid\ \forall i=1\dots n\ ]\,,\\
    L &= [\ \argmin_{B_j,\ j = 1\dots n} {\|B_j - P_i\|}^2\ \mid\ \forall i=1\dots n\ ]\,.
  \end{align*}
  Section~\ref{sec:rep} outlined the use of a constraint vector for describing the feasibility of a solution. This
  concept generalizes to a matrix $C$, where,
  \begin{equation*}
    C = [\ c_{ij}\ \mid\ \forall i=1\dots n,\;\forall j=1\dots s\ ]\,.
  \end{equation*}
  Let $r',r''$ be two random vectors of length $n$ drawn from the uniform distribution ${U(0,1)}^n$. We use the
  shorthand notation of $X_i$ to denote the $i$\tsup{th} row of some matrix $X$. If $f(X_i)$ is the fitness function and
  $\mathit{constraints}(X)$ constructs $C$ for solutions $X$, the implementation of each iteration while updating
  particles in Algorithm~\ref{alg:cop} reduces to,

  \begin{align*}
    V  =&\ \omega.V + \lambda_g\begin{bmatrix}
      {r_1}'(\gbest - P_1)\\
      {r_2}'(\gbest - P_2)\\
      \vdots\\
      {r_n}'(\gbest - P_n)
    \end{bmatrix} + \lambda_p\begin{bmatrix}
      {r_1}''(L_1 - P_1)\\
      {r_2}''(L_2 - P_2)\\
      \vdots\\
      {r_n}''(L_n - P_n)
    \end{bmatrix}\\
    V  =&\ \left[\ \sgn(v_{ij})*\max\{|v_{max}|,|v_{ij}|\}\ \mid\ \forall i=1\dots n,\;\forall j=1\dots d\ \right]\\
    P  =&\ P + V\\
    C  =&\ \mathit{constraints}(P)\\
    B  =&\ [\ X_i\ \mid\ (\textstyle\sum_{j=1}^s c_{ij} = 0 \to X_i = \argmax\{f(P_i), f(B_i)\})\ \wedge\\
        &\qquad(\textstyle\sum_{j=1}^s c_{ij} \neq 0 \to X_i = B_i),\ \forall i=1\dots n\ ]\\
    \gbest =&\ \argmax_{{B}_i} \{\ {f(B_i)}\ \mid\ \forall i=\dots n\ \}
  \end{align*}

>>>>

The search for extra-terrestrial life~\cite{Seth1,Biosig} and potentially habitable extrasolar
planets~\cite{Kepler,Presto} has been an international venture since Frank Drake's attempt with Project Ozma in the
mid-20th century~\cite{Ozma}. \textcite{Exo1} confirmed the first exoplanet in 1991. This marked the start of a trend
that has lasted 25 years and yielded over 3,700 confirmed exoplanets. There have been attempts to assess the
habitability of these planets and to assign a score based on their similarity to Earth. Two such habitability scores are
the Cobb-Douglas Habitability Score (CDHS)~\cite{Bora,Saha} and the Constant Elasticity Earth Similarity Approach
(CEESA) score. Estimating these scores involves maximizing a production function while observing a set of constraints on
the input variables.

Under most paradigms, maximizing a continuous function requires calculating a gradient. This may not always be feasible
for non-polynomial functions in high-dimensional search spaces. Further, subjecting the input variables to constraints,
as needed by CDHS and CEESA, are not always straightforward to represent within the model. This paper details an
approach to Constrained Optimization (CO) using the swarm intelligence metaheuristic. \pso\ (PSO) is a method for
optimizing a continuous function that does away with the need for calculating the gradient. It employs a large number of
randomly initialized particles that traverse the search space, eventually converging at a global best solution
encountered by at least one particle~\cite{PSO1,PSO2}.

\pso\ is a distributed method that requires simple mathematical operators and short segments of code, making it a
lucrative solution where computational resources are at a premium. Its implementation is highly parallelizable. It
scales with the dimensionality of the search space. The standard PSO algorithm does not deal with constraints but,
through variations in initializing and updating particles, constraints are straightforward to represent and adhere to,
as seen in Section~\ref{subsec:copso}. \textcite{Poli1,Poli2} carried out extensive surveys on the applications of PSO,
reporting uses in Communication Networks, Machine Learning, Design, Combinatorial Optimization and Modeling, among
others.

This paper demonstrates the applicability of \pso\ in estimating the CDHS and CEESA scores of an exoplanet by maximizing
their respective production functions (discussed in Sections~\ref{subsec:cdhs} and~\ref{subsec:ceesa}). CDHS considers
the planet's Radius, Mass, Escape Velocity and Surface Temperature, while CEESA includes a fifth parameter, the Orbital
Eccentricity of the planet. The Exoplanets Catalog hosted by the Planetary Habitability Laboratory, UPR Arecibo records
these parameters for each exoplanet in Earth Units~\cite{PHL}. Section~\ref{sec:results} reports the performance of PSO
and discusses the distribution of the habitability scores of the exoplanets.

@@

\subsection{Cobb-Douglas Habitability Score}\label{subsec:cdhs}
Estimating the Cobb-Douglas Habitability Score (CDHS)~\cite{Bora} requires estimating an interior score \linebreak
($\mathit{CDHS}_i$) and a surface score ($\mathit{CDHS}_s$) by maximizing the following production functions,
\begin{subequations}
  \begin{alignat}{4}
    Y_i\ &=\ {CDHS}_i\ &=&\ R^\alpha.&D^\beta\,,\label{eq:cdhsi}\\
    Y_s\ &=\ {CDHS}_s\ &=&\ {V_e}^\gamma.&{T_s}^\delta\,,\label{eq:cdhss}
  \end{alignat}
\end{subequations}
where, $R$, $D$, $V_e$ and $T_s$ are density, radius, escape velocity and surface temperature respectively. $\alpha$,
$\beta$, $\gamma$ and $\delta$ are the elasticity coefficients subject to $0 < \alpha,\beta,\gamma,\delta < 1$.
Equations~\ref{eq:cdhsi} and~\ref{eq:cdhss} are convex under either Constant Returns to Scale (CRS), when
$\alpha+\beta=1$ and $\gamma+\delta=1$, or Decreasing Returns to Scale (DRS), when $\alpha+\beta<1$ and
$\gamma+\delta<1$. The final CDHS is the convex combination of the interior and surface scores as given by,
\begin{equation}
  Y\ =\ w_i.Y_i + w_s.Y_s\,.
\end{equation}

\subsection{Constant Elasticity Earth Similarity Approach}\label{subsec:ceesa}
% Cite what?
The Constant Elasticity Earth Similarity Approach (CEESA) uses the following production function to estimate the
habitability score of an exoplanet,
\begin{equation}\label{eq:ceesa}
  Y = {(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{\eta}{\rho}}\,,
\end{equation}
where, $E$ is the fifth parameter denoting Orbital Eccentricity. The value of $\rho$ lies within $0<\rho\leq 1$. The
coefficients $r$, $d$, $t$, $v$ and $e$ lie in $(0,1)$ and sum to \num{1}, $r+d+t+v+e = 1$. The value of $\eta$ is
constrained by the scale of production used, $0 < \eta < 1$ under DRS and $\eta=1$ under CRS.

@@

Particle Swarm Optimization (PSO)~\cite{PSO1} is a biologically inspired metaheuristic for finding the global minima of
a function. Traditionally designed for unconstrained inputs, it works by iteratively converging a population of randomly
initialized solutions, called particles, toward a globally optimal solution. Each particle in the population keeps track
of its current position and the best solution it has encountered, called $\pbest$. Each particle also has an associated
velocity used to traverse the search space. The swarm keeps track of the overall best solution, called $\gbest$. Each
iteration of the swarm updates the velocity of the particle towards its $\pbest$ and the $\gbest$ values.

@@

On each iteration, the algorithm updates the velocity and position of each particle. For each particle, it picks two
random numbers $u_g, u_p$ from a uniform distribution, $U(0,1)$ and updates the particle velocity as indicated in
line~\ref{algline:vup}. Here, $\omega$ is the inertial weight and $\lambda_g,\lambda_p$ are the global and particle
learning rates. If the new position of the particle corresponds to a better fit than its $\pbest$, the algorithm updates
$\pbest$ to the new position. Once the algorithm has updated all particles, it updates $\gbest$ to the new overall best
position. A suitable termination criteria for the swarm, under convex optimization, is to terminate when $\gbest$ has
not changed by the end of an iteration.

@@

Under these guidelines, the representation of CDHS estimation (described in Section~\ref{subsec:cdhs}) under CRS as a
CO problem is given by,

\begin{equation}\label{eq:cdhscrs}
  \begin{aligned}
    & \underset{\alpha,\beta,\gamma,\delta}{\text{minimize:}}
    & Y_i = {-}R^\alpha.D^\beta,\ Y_s = {-}{V_e}^\gamma.{T_s}^\delta\\
    & \text{subject to:}
    &  {-}\phi + \epsilon \leq 0,\quad\phi - 1 + \epsilon \leq 0\,, &\quad\quad \forall \phi\in\{\alpha,\beta,\gamma,\delta\}\\
    && (\alpha+\beta-1) - \tau \leq 0,\quad(\gamma+\delta-1) - \tau \leq 0\,,\\
    && (1-\alpha-\beta) - \tau \leq 0,\quad(1-\gamma-\delta) - \tau \leq 0\,.
  \end{aligned}
\end{equation}
Under DRS the last two constraints for $Y_i$ and $Y_s$ are replaced with,
\begin{equation}\label{eq:cdhsdrs}
  \begin{aligned}
    \alpha + \beta + \epsilon - 1 &\leq 0\,,\\
    \gamma + \delta + \epsilon - 1 &\leq 0\,.
  \end{aligned}
\end{equation}

The representation of CEESA score estimation (described in Section~\ref{subsec:ceesa}) under DRS as a CO problem is
given by,
\begin{equation}\label{eq:ceesadrs}
  \begin{aligned}
    & \underset{r,d,t,v,e,\rho,\eta}{\text{minimize}}
    & Y = {-}{(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{\eta}{\rho}}\\
    & \text{subject to}
    &   {-}\phi + \epsilon \leq 0,\quad\phi - 1 + \epsilon \leq 0\,, &\quad\quad \forall \phi\in\{r,d,t,v,e,\eta\}\\
    && \rho - 1 \leq 0,\quad \rho - 1 + \epsilon \leq 0\,,\\
    && (r+d+t+v+e-1) - \tau \leq 0\,,\\
    && (1-r-d-t-v-e) - \tau \leq 0\,.
  \end{aligned}
\end{equation}
Under CRS there is no need for the parameter $\eta$ (since $\eta=1$). Thus, the objective function for the problem reduces to,
\begin{equation}\label{eq:ceesacrs}
  \underset{r,d,t,v,e,\rho}{\text{minimize}}\quad
  Y = {-}{(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\frac{1}{\rho}}\,.
\end{equation}

@@

The data set used for estimating the Habitability Scores of exoplanets was the Confirmed Exoplanets Catalog maintained
by the Planetary Habitability Laboratory (PHL)~\cite{PHL}. The catalog records observed and modeled parameters for
exoplanets confirmed by the Extrasolar Planets Encyclopedia. Table~\ref{tab:param} describes the parameters from the PHL
Exoplanets' Catalog (PHL-EC) used for the experiment. Since surface temperature and eccentricity are not recorded in
Earth Units, we normalized these values by dividing them with Earth's surface temperature (\SI{288}{\kelvin}) and
eccentricity (\num{0.017}). PHL-EC assumes an Eccentricity of 0 when unavailable. The PHL-EC records empty values for
planets whose surface temperature is not known. We chose to drop these records from the experiment.

Our implementation used $n=25$ particles to traverse the search space, with learning rates $\lambda_g=0.8$ and
$\lambda_p=0.2$. It used an inertial weight of $\omega=0.6$ and upper and lower bounds $\pm1.0$. We used an error
threshold of $\epsilon=\num{1e-6}$ to convert strict inequalities to non-strict inequalities, and a tolerance of
$\tau=\num{1e-7}$ to transform an equality constraint to a pair of inequalities. Further implementation details are
discussed in the Appendix.

@@

As for real time taken to converge, PSO took \SI{666.85}{\second} ($\approx\SI{11}{\minute}\ \SI{7}{\second}$) to
estimate the CDHS under CRS for \num{1683} exoplanets, at an average of \SI{198.11}{\milli\second} for each planet for
each individual score (interior and surface) of the CDHS.\@ For CDHS under DRS, it took \SI{638.69}{\second}
($\approx\SI{10}{\minute}\ \SI{39}{\second}$) at an average of \SI{189.75}{\milli\second} for each part of the CDHS.\@
The CEESA calculations, requiring a single estimate, took a little over half the CDHS execution time to run. Under DRS
it took a total of \SI{370.86}{\second} ($\approx\SI{6}{\minute}\ \SI{11}{\second}$) at \SI{220.36}{\milli\second} per
planet, while under CRS it took \SI{356.92}{\second} ($\approx\SI{5}{\minute}\ \SI{57}{\second}$) at
\SI{212.07}{\milli\second} per planet.

@@

\pso\ mainly draws its advantages from being easy to implement and highly parallelizable. The algorithms described in
Section~\ref{sec:pso} use simple operators and straightforward logic. What is especially noticeable is the lack of the
need for a gradient, allowing PSO to work in high dimensional search spaces with a large number of constraints,
precisely what is needed in a potential Habitability score estimate. Further, particles of the swarm, in most
implementations operate independently during each iteration, their updates can occur simultaneously and even
asynchronously, yielding much faster execution times than those outlined in Section~\ref{sec:results}. However, since
strict inequalities and equality constraints are not exactly represented, the resulting solution may not be as accurate
as direct methods. Despite this, using PSO to calculate the habitability scores is beneficial when the number of input
parameters are large, which further increases the number of constraints, resulting in a model too infeasible for
traditional optimization methods.

@@

  \section{Ensuring Convergence}\label{app:con}

@@

  To prevent this from happening our implementation maintained a counter that it incremented every time $\gbest$ changed
  by a value less than the threshold. When the counter reached \num{100}, the algorithm terminated and reported the
  current value of $\gbest$ as the global optima. If the value of $\gbest$ changed by more than the threshold, it reset
  the counter to $\num{0}$. Since the function is convex, and all points are initialized to feasible solutions (thus,
  ensuring that a feasible $\gbest$ has been encountered) there was no need to restrict the total number of possible
  iterations.

  This explains the spike at \num{0} in Figure~\ref{fig:itercdcrs}. For every exoplanet, the implementation waits a
  \num{100} iterations for declaring convergence, thus overstating the total iterations by \num{100}. To adjust for this
  offset, we subtracted \num{100} from the total iterations before constructing the histogram plots in
  Figures~\ref{fig:itercdcrs},~\ref{fig:itercddrs},~\ref{fig:itercecrs} and~\ref{fig:itercedrs}. Therefore, although the
  graphs report iterations to convergence for a significant number of planets, only $\gbest$ has not shifted, the
  particles have actually converged to a common point around a region within the threshold of the global best.

@@

  Matrices $P$ and $V$ represent current position and velocity of all particles in the swarm.
  \begin{alignat*}{2}
    P &= [\ p_{ij}\ &&\mid\ \forall i=1\dots n,\;\forall j=1\dots d\ ]\,,\\
    V &= [\ v_{ij}\ &&\mid\ \forall i=1\dots n,\;\forall j=1\dots d\ ]\,.
  \end{alignat*}
  $p_{ij}$ and $v_{ij}$ denote position and velocity of particle $i$ in dimension $d$. Matrices $B$ and $L$ represent
  the $\pbest$ and $\lbest$ values for the swarm, where,
  \begin{align*}
    B &= [\ {\pbest}_i\ \mid\ \forall i=1\dots n\ ]\,,\\
    L &= [\ \argmin_{B_j,\ j = 1\dots n} {\|B_j - P_i\|}^2\ \mid\ \forall i=1\dots n\ ]\,.
  \end{align*}
  Section~\ref{sec:rep} outlined the use of a constraint vector for describing the feasibility of a solution. This
  concept generalizes to a matrix $C$, where,
  \begin{equation*}
    C = [\ c_{ij}\ \mid\ \forall i=1\dots n,\;\forall j=1\dots s\ ]\,.
  \end{equation*}
  Let $r',r''$ be two random vectors of length $n$ drawn from the uniform distribution ${U(0,1)}^n$. We use the
  shorthand notation of $X_i$ to denote the $i$\tsup{th} row of some matrix $X$. If $f(X_i)$ is the fitness function and
  $\mathit{constraints}(X)$ constructs $C$ for solutions $X$, the update of each particle in Algorithm~\ref{alg:cop} is
  implemented as,
  \begin{align*}
    V  =&\ \omega.V + \lambda_g\begin{bmatrix}
      {r_1}'(\gbest - P_1)\\
      {r_2}'(\gbest - P_2)\\
      \vdots\\
      {r_n}'(\gbest - P_n)
    \end{bmatrix} + \lambda_p\begin{bmatrix}
      {r_1}''(L_1 - P_1)\\
      {r_2}''(L_2 - P_2)\\
      \vdots\\
      {r_n}''(L_n - P_n)
    \end{bmatrix}\\
    V  =&\ \left[\ \sgn(v_{ij})*\max\{|v_{max}|,|v_{ij}|\}\ \mid\ \forall i=1\dots n,\;\forall j=1\dots d\ \right]\\
    P  =&\ P + V\\
    C  =&\ \mathit{constraints}(P)\\
    B  =&\ [\ X_i\ \mid\ (\textstyle\sum_{j=1}^s c_{ij} = 0 \to X_i = \argmax\{f(P_i), f(B_i)\})\ \wedge\\
        &\qquad(\textstyle\sum_{j=1}^s c_{ij} \neq 0 \to X_i = B_i),\ \forall i=1\dots n\ ]\\
  \end{align*}
  Finally, after all particles are updated, $\gbest$ is updated according to,
  \begin{equation*}
    \gbest = \argmax_{{B}_i} \{\ {f(B_i)}\ \mid\ \forall i=\dots n\ \}\,.
  \end{equation*}
